---
title: "500 Final Project"
author: "Hiren Jethra, Andres Sanches, Peter, Abhishek"
date: "2023-02-04"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
#install.packages("caTools")

library("ggplot2")
library("caTools")
library("corrplot")
library("dplyr")

url <- "https://ftp.cs.wisc.edu/math-prog/cpo-dataset/machine-learn/cancer/WDBC/WDBC.dat"
data <- read.csv(url)
head(data)
```

# Clean the Data
```{r}

data <- read.csv(file = url, header = FALSE,
                col.names = c("id","diagnosis", "radius_mean", "texture_mean", "perimeter_mean", "area_mean", "smoothness_mean", "compactness_mean", "concavity_mean","concave.points_mean", "symmetry_mean", "fractal_dimension_mean", "radius_se", "texture_se", "perimeter_se", "area_se", "smoothness_se", "compactness_se", "concavity_se", "concave.points_se", "symmetry_se", "fractal_dimension_se", "radius_worst","texture_worst","perimeter_worst", "area_worst","smoothness_worst","compactness_worst","concavity_worst","concave.points_worst","symmetry_worst","fractal_dimension_worst") )

#structure of the dataset
str(data)

#dimension of data set
dim(data)

#summary of data set
summary(data)

#Data checked, no inconsistent values. No values removed. 

```

#Count of patients by diagnosis stage

```{r}
data %>% count(diagnosis)
```

#Percentage of patients by diagnosis stage
```{r}
data %>% count(diagnosis)%>%group_by(diagnosis) %>%
  summarize(perc_dx = round((n / 569)* 100, 2))
```

#DATA VISUALIZATION
#Percentage of patients by diagnosis stage
```{r}
diagnosis.table <- table(data$diagnosis)
colors <- rainbow(2,alpha=0.4) 

# Pie chart 
diagnosis.prop.table <- prop.table(diagnosis.table)*100
diagnosis.prop.df <- as.data.frame(diagnosis.prop.table)
pielabels <- sprintf("%s - %3.1f%s", diagnosis.prop.df[,1], diagnosis.prop.table, "%")
pie(diagnosis.prop.table,
  labels=pielabels,  
  clockwise=TRUE,
  col=colors,
  border="gainsboro",
  radius=0.8,
  cex=0.8, 
  main="Frequency of Cancer Diagnosis")
legend(1, .4, legend=diagnosis.prop.df[,1], cex = 0.7, fill = colors)
```
#Correlation Plot

#We are interested in knowing how the independent variables relate to each other. We do that by looking at correlations between them.

```{r}
# calculate collinearity
c <- cor(data[,3:12])
corrplot(c, order = "hclust", tl.cex = 0.7)
```
#Analyzing box plots for radius,area and concavity by diagnosis stage

```{r}
ggplot(data=data,aes(x=diagnosis,y=radius_mean))+geom_boxplot(color="red", fill="orange", alpha=0.2)+ggtitle("Radius of Benign Vs Malignant")
```
```{r}
ggplot(data=data,aes(x=diagnosis,y=area_mean))+geom_boxplot(color="green", fill="yellow", alpha=0.2)+ggtitle("Area of Benign Vs Malignant")
```
```{r}
ggplot(data=data,aes(x=diagnosis,y=concavity_mean))+geom_boxplot(color="blue", fill="purple", alpha=0.2)+ggtitle("Concavity of  Benign Vs Malignant")
```
#As shown by analysis, Malignant cells have higher radius,area and concavity mean than Benign cells. 

#Density plot based on texture mean

```{r}
ggplot(data,aes(x=texture_mean,fill=as.factor(diagnosis)))+geom_density(alpha=0.2)+ggtitle("Texture mean  for Benign vs Malignant")

```
Texture mean is higher for Malignant cells. 

#Analysing perimeter mean by Diagnosis stage

```{r}
ggplot(data, aes(x = perimeter_mean)) +
  geom_histogram(binwidth=10, color="darkblue",fill="Pink", alpha=0.5, ) +
  facet_wrap(~ diagnosis)+
  ggtitle("Perimeter mean  for benign vs malignant")
```
Analyzing concavity mean for benign and malignant cells

```{r}
#data1=data%>%filter(concavity_mean>0)
ggplot(data,aes(x=concavity_mean,y=diagnosis))+
  geom_point(color="purple",fill="yellow", alpha=0.5)+
  ggtitle("Concavity mean  for benign vs malignant")
```
# We would like to predict the diagnosis stage using a logistic regression model. Since our outcome/dependent variable is binary/categorical, using logistic regression with binomial family of distribution makes most sense.

```{r}
# First we correct the data type of our dependent variable to be factor
data$diagnosis<-factor(data$diagnosis,levels=c("B","M"),labels=c(0,1))

#Setting seed so that the analysis can be repeated and have same results
set.seed(100)

#Splitting the data into train and test using 75/25 ratio

split=sample.split(data$diagnosis,SplitRatio=0.75)

#Filtering the data to only the mean variables and then splitting into train & test
training_set<-subset(data[,2:12],split==T)
head(training_set)
test_set<-subset(data[,2:12],split==F)
head(test_set)
```

```{r}
# We are normalizing the variables so that they can become comparable and used in same model

training_set[,2:11]<-scale(training_set[,2:11])
head(training_set)
test_set[,2:11]<-scale(test_set[,2:11])
head(test_set)
```


```{r}
# Logistic regression model
log_reg = glm(diagnosis ~ ., data=training_set, family=binomial)
summary(log_reg)

# We can see that texture mean, concavity mean, concave.points_mean and symmetry mean are significant.gj 

#Hypothesis test:
#H0: Texture mean is not significant in predicting diagnosis stage
#Ha: Texture mean is significant in predicting diagnosis stage

# As we see texture mean has p-value<0.05, we can reject null hypothesis and say that we have evidence that texture mean is significant in predicting diagnosis stage.
```

```{r}
# Model predicts the probability of diagnosis stage = malignant, on the test data set
predict_proba<-predict(object=log_reg,type="response",newdata=test_set[-1])
head(predict_proba)
```

```{r}
# Recreating predicted diagnosis variable by logic: if probability>0.5 then 1 else 0
y_pred<-ifelse(predict_proba>0.5,1,0)
head(y_pred)
```

```{r}
# Evaluating the model by looking at the confusion matrix - Comparison between actual vs predicted diagnosis stage in the test dataset
confusion_matrix<-table(test_set[,1],y_pred)
confusion_matrix
```

```{r}
# Calculating accuracy by: (True Positives + True Negatives/Total Observations)
accuracy<-sum(diag(confusion_matrix))/sum(confusion_matrix) 
accuracy
```

